---
title: "2D Timeseries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2D Timeseries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Define parameters for cropping and subsetting
```{r setup}
library(here)
library(EDABUtilities)

epu.shp =system.file('data','EPU_NOESTUARIES.shp',package = 'EDABUtilities')
var.name = 'BottomS'
min.S = 30
max.S = 34

```

Crop the data based on the lat/lon bounds of epu.shp
```{r}

data.crop = crop_nc_2d(
  data.in = c(system.file('data','GLORYS_daily_BottomSalinity_2019.nc',package = 'EDABUtilities'),
                  system.file('data','GLORYS_daily_BottomSalinity_2020.nc',package = 'EDABUtilities')),
  shp.file = epu.shp,
  var.name = var.name,
  write.out = F,
  area.names =NA
)

terra::plot(data.crop[[2]])
```
Calculate the number of days above a reference salinity
```{r}
data.nday =make_2d_deg_day_ts(
  data.in = data.crop,
  shp.file = epu.shp,
  area.names =  c('MAB','GB'),
  var.name = 'BottomS',
  statistic = 'nd',
  ref.value = 33,
  type = 'above',
  write.out = F)

data.nday.all = dplyr::bind_rows(data.nday)
ggplot2::ggplot(data.nday.all,ggplot2::aes(x = ls.id, y = value, fill = area))+
  ggplot2::geom_bar(stat = 'identity', position = 'dodge')
  
```

Filter values within min.S to max.S range and remove values outside of the epu.shp regions
```{R}

data.mask = mask_nc_2d(
  data.in = data.crop,
  shp.file = epu.shp,
  min.value = min.S,
  max.value = max.S,
  binary = F,
  area.names = NA
)

terra::plot(data.mask[[1]])
```

Convert data.mask to data.frame aggregated by month and grouped by the regions MAB & GB

```{R}
data.ts = make_2d_summary_ts(data.in = data.mask,
                           shp.file = epu.shp,
                           var.name = var.name,
                           agg.time ='months',
                           statistic = 'mean',
                           area.names = c('MAB','GB'),
                           write.out = F)

ggplot2::ggplot(data.ts[[1]],ggplot2::aes(x = time, y= value))+
  ggplot2::geom_line()+
  ggplot2::facet_wrap(~area)

```

Test creating the same function without a for loop
```{r}
make_2d_summary_ts2 = function(data.in,
                              output.files,
                              shp.file,
                              area.names,
                              var.name,
                              agg.time,
                              tz = NA,
                              statistic,
                              touches =TRUE,
                              write.out = FALSE){
  
  if(!is.na(shp.file)){
    shp.vect = terra::vect(shp.file)
  }
  
   if(is.character(data.in)){
      
      datab = terra::rast(data.in)
      
    }else if(class(data.in[[1]])[1] == 'SpatRaster'){
      
      datab = data.in
      
    }else{
      stop('data.in needs to be either file names or spatRasters')
    } 
  
  out.ls = list()
  for(i in 1:length(datab)){
    
   data <- datab[i]
    
    data.time = as.Date(terra::time(data))
    if(!is.na(tz)){
      data.time = as.Date(as.POSIXct(data.time,tz = tz),tz = tz)
      terra::time(data) = data.time
    }
    
    if(agg.time == 'season'){
      month.season = data.frame(month=1:12,season =rep(1:4,each =3))
      data.month = as.numeric(format(data.time,format = "%m"))
      data.season = month.season$season[data.month]
      season.names = 1:4
    }
    
    
    if(!is.na(shp.file)){
      
      shp.str = as.data.frame(shp.vect)
      which.att = which(apply(shp.str,2,function(x) all(area.names %in% x)))
      which.area =  match(area.names,shp.str[,which.att])
      
      all_area <- purrr::map2(.x = which.area, .y = area.names,
                             .f = function(.x, .y, 
                                            data2 = data,
                                           shp.vect2 = shp.vect,
                                           touches2 = touches,
                                           agg.time2 = agg.time,
                                           statistic2 = statistic,
                                           season.names,
                                           var.name = "test",
                                           data.in = datab,
                                           k = i) {
                                 area.data = terra::mask(data2,shp.vect2[.x,], 
                                                         touches = touches2)
                                
                                  area.agg = terra::tapp(area.data,
                                                        fun = statistic2,
                                                        index = ifelse(agg.time2 == "season", 
                                                                       data.season, 
                                                                       agg.time2))
                                
                                  time.out = ifelse(agg.time2 == "season", 
                                                   season.names, 
                                                   terra::time(area.agg))
                                 
                                 area.stat = terra::global(area.agg,statistic2,na.rm=TRUE)
        
                                 data.frame(time = time.out,
                                            agg.time = agg.time,
                                            ls.id = ifelse(is.character(data.in),data.in[k],k),
                                            var.name = var.name,
                                            statistic = statistic2,
                                            area = .y,
                                            value =area.stat[,1])
                             })
      data.stat.df = dplyr::bind_rows(all_area)

    }else{

      data.stat = terra::global(terra::tapp(data, 
                                           fun = statistic,
                                           index = ifelse(agg.time == "season",
                                                          data.season, 
                                                          agg.time)),
                                statistic,
                                na.rm=TRUE)
      
      data.stat.df = data.frame(time = ifelse(agg.time == "season", 
                                              season.names, 
                                              terra::time(data.agg)),
                                agg.time =agg.time,
                                ls.id = ifelse(is.character(data.in),data.in[i],i),
                                var.name = var.name,
                                statistic = statistic,
                                area = NA,
                                value =data.stat[,1])
    }
    
    if(write.out){
      saveRDS(data.stat.df, output.files[i])
    }else{
      out.ls[[i]] = data.stat.df
    }

  }
  
  if(write.out ==F){
    return(out.ls)  
  }
  
}

```

Microbenchmark
```{r}
dat <- microbenchmark::microbenchmark(
  make_2d_summary_ts(data.in = data.mask,
                           shp.file = epu.shp,
                           var.name = var.name,
                           agg.time ='months',
                           statistic = 'mean',
                           area.names = c('MAB','GB'),
                           write.out = F),
  make_2d_summary_ts2(data.in = data.mask,
                           shp.file = epu.shp,
                           var.name = var.name,
                           agg.time ='months',
                           statistic = 'mean',
                           area.names = c('MAB','GB'),
                           write.out = F),
  times = 20
)

dat
```


```{r}
lil_fxn <- function(data2 = data,
                    shp.vect2 = terra::vect(epu.shp),
                    # agg.time2 = agg.time,
                    statistic2 = "mean",
                    # season.names,
                    var.name = "test"#,
                    # data.in = datab,
                    # k = i
                    ) {
                                 area.data = terra::mask(data2,
                                                         shp.vect2, 
                                                         touches = TRUE)
                                
                                  area.agg = terra::tapp(area.data,
                                                        fun = statistic2,
                                                        index = "months")
                                
                                  time.out = terra::time(area.agg)
                                 
                                 area.stat = terra::global(area.agg,
                                                           statistic2,
                                                           na.rm=TRUE)
        
                                 out <- data.frame(time = time.out,
                                            agg.time = "months",
                                            # ls.id = ifelse(is.character(data.in),data.in[k],k),
                                            var.name = var.name,
                                            statistic = statistic2,
                                            # area = .y,
                                            value =area.stat[,1]) 
                                 
                                 return(out)
}



tst <- lil_fxn(data2 = data.mask[[1]], shp.vect2 = terra::vect(epu.shp)[1,])

test_dat <- rep(data.mask, 10)
shape <- terra::vect(epu.shp)[c(1,4),]

output <- microbenchmark::microbenchmark(
  {out <- list()
  for (i in 1:length(test_dat)){
    out[[i]] <- lil_fxn(data2 = test_dat[[i]], shp.vect2 = terra::vect(epu.shp)[1,])
    }
  data.stat.df = dplyr::bind_rows(out)
  },
  purrr::map(test_dat, lil_fxn),
  times = 10)

output <- microbenchmark::microbenchmark(
  {out <- list()
  for (i in 1:length(test_dat)){
    for (j in 1:length(shape)) {
      out <- list(out, 
                  lil_fxn(data2 = test_dat[[i]], shp.vect2 = terra::vect(epu.shp)[j,]))
    }}
    out
    },
  purrr::map2(test_dat, shape, lil_fxn(data2 = .x, shp.vect2 = .y)),
  times = 10)


output
```
