---
title: "2D Gridded"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2D Gridded}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Define parameters for cropping and subsetting
```{r setup}
library(here)
library(EDABUtilities)

epu.shp =system.file('data','EPU_NOESTUARIES.shp',package = 'EDABUtilities')
var.name = 'BottomS'
min.S = 30
max.S = 34

```

Crop the data based on the lat/lon bounds of epu.shp
```{r}

data.crop = crop_nc_2d(
  data.in = c(system.file('data','GLORYS_daily_BottomSalinity_2019.nc',package = 'EDABUtilities'),
                  system.file('data','GLORYS_daily_BottomSalinity_2020.nc',package = 'EDABUtilities')),
  shp.file = epu.shp,
  var.name = var.name,
  write.out = F,
  area.names =NA
)

terra::plot(data.crop[[1]])
```
Filter values within min.S to max.S range and remove values outside of the epu.shp regions
```{R}

data.mask = mask_nc_2d(
  data.in = data.crop,
  shp.file = epu.shp,
  min.value = min.S,
  max.value = max.S,
  binary = F,
  area.names = NA
)

terra::plot(data.mask[[1]])
```
Create a summary spatRaster with monthly mean salinity for GB and MAB

```{R}
data.summary =make_2d_summary_gridded(data.in = data.mask,
                      write.out = F,
                      shp.file = epu.shp,
                      var.name = var.name,
                      agg.time = 'months',
                      statistic = 'mean',
                      area.names = c('MAB','GB'))


terra::plot(data.summary[[1]])
```
Create a monthly climatology from the raw gridded data. Filter data by summer months (6 through 8)
``` {R}
data.clim = make_2d_climatology_gridded(data.in = data.mask,
                                          write.out = F,
                                          shp.file = epu.shp,
                                          var.name = var.name,
                                          area.names = c('MAB','GB'),
                                          start.time = 6, 
                                          stop.time =8,
                                          agg.time = 'months',
                                          statistic = 'mean'
                                          )
  
terra::plot(data.clim)
```

Calculate the anomaly from original ts data and climatology
```{R}
data.anom =make_2d_anomaly_gridded(data.in = data.summary,
                                   climatology = data.clim,
                                   shp.file = epu.shp,
                                   var.name = var.name,
                                   area.names = c('MAB','GB'),
                                   write.out =F)

terra::plot(data.anom[[1]])
```
Calculate number of days above reference temperature
```{R}
data.nd = make_2d_deg_day_gridded_nc(
  data.in = data.mask,
  write.out = F,
  shp.file = epu.shp,
  var.name = 'BottomS',
  statistic = 'nd',
  ref.value = 33,
  type = 'above',
  area.names = c('MAB','GB')
)

terra::plot(data.nd[[1]])
```