---
title: "2D Gridded"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2D Gridded}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)

```

# Pipeline Workflow: Gridded

The `EDABUtilities` package provides a robust, spatially-explicit pipeline to process 2D gridded datasets (like multi-layer NetCDF files). Unlike the Timeseries Pipeline (`_ts`), which summarizes spatial zones into tabular data frames, the Gridded Pipeline (`_gridded`) preserves the spatial dimensions of your data throughout the entire analytical workflow.

Every operation along this pipeline consumes a `SpatRaster` (or NetCDF file path) and returns a modified `SpatRaster` (or writes a NetCDF directly to disk). This is essential for workflows where you need to map outputs, calculate cell-by-cell metrics, or feed downstream spatial models.

This vignette demonstrates a 6-step progression:

1. *Spatial Cropping* (`crop_nc_2d`): Limit the spatial boundaries of the input layers to the bounding box of your shapefile.

2. *Quality Masking* (`mask_nc_2d`): Clamp data values to a specific quality threshold and mask pixels outside your study area.

3. *Temporal Aggregation* (`make_2d_summary_gridded`): Aggregate daily layers into weekly, monthly, seasonal, or annual gridded summaries.

4. *Gridded Climatology* (`make_2d_climatology_gridded`): Combine multiple years of gridded summaries to construct a baseline spatial climatology.

5. *Gridded Anomaly Calculation* (`make_2d_anomaly_gridded`): Subtract the spatial baseline climatology from your gridded summaries to yield a map of cell-by-cell anomalies.

6. *Gridded Degree Day Calculation* (`make_2d_degree_day_gridded`): Calculate the number of days above or below a reference threshold for each cell in your gridded dataset.

## Step 0: Setup and Define Parameters

First, we load the required packages and define our core parameters, including our study area shapefile (Ecological Production Units - EPUs) and our variable of interest (Bottom Salinity).

```{r setup}
library(here)
library(EDABUtilities)
library(terra)

# Define reference boundaries and thresholds
epu.shp <- system.file("data", "EPU_NOESTUARIES.shp", package = "EDABUtilities")
var.name <- "BottomS"
min.S <- 30
max.S <- 34

```

## Step 1: Spatial Cropping

We crop the multi-layer input NetCDF files down to the bounding box of the shapefile. This reduces the footprint of the spatial grids in memory and increases processing speeds for all downstream steps.

```{r}
data.crop <- crop_nc_2d(
  data.in = c(
    system.file("data", "GLORYS_daily_BottomSalinity_2019.nc", package = "EDABUtilities"),
    system.file("data", "GLORYS_daily_BottomSalinity_2020.nc", package = "EDABUtilities")
  ),
  shp.file = epu.shp,
  var.name = var.name,
  area.names = NA,
  write.out = FALSE
)

# Plot the cropped daily spatial layer
terra::plot(data.crop[[1]], main = "Cropped Gridded Extent (Layer 1)")
```

## Step 2: Masking and Quality Control

Next, we filter out any values that fall outside our acceptable bottom salinity threshold (30 to 34 PSU) and set pixels that reside outside of the precise EPU polygon boundary to `NA`.

```{R}

data.mask <- mask_nc_2d(
  data.in = data.crop,
  var.name = var.name,
  min.value = min.S,
  max.value = max.S,
  shp.file = epu.shp,
  binary = FALSE,
  area.names = NA,
  write.out = FALSE
)

# Plot the masked result
terra::plot(data.mask[[1]], main = "Masked and Thresholded Grid")
```

## Step 3: Temporal Aggregation to Gridded Summaries

Instead of converting to tabular forms, we use `make_2d_summary_gridded to aggregate` our daily timesteps into monthly spatial averages. We filter the results specifically to the Georges Bank (GB) and Mid-Atlantic Bight (MAB) regions.

```{R}
data.summary <- make_2d_summary_gridded(
  data.in = data.mask,
  write.out = FALSE,
  file.time = "annual",
  shp.file = epu.shp,
  var.name = var.name,
  agg.time = "months",
  statistics = "mean",
  area.names = c("MAB", "GB")
)

#Pull just the top layer from data.summary for plotting
plot.data = terra::rast(data.summary[[1]])

# Plot the monthly gridded summary
terra::plot(plot.data, main = "Monthly Mean Salinity Grid")
```

## Step 4: Construct Gridded Climatology Baseline

Now we establish a spatial baseline reference. By using `make_2d_climatology_gridded`, we calculate the cell-by-cell average specifically for the summer months (June through August) across the multi-year input layers. This returns a single climatological `SpatRaster` layer.

``` {R}
data.clim <- make_2d_climatology_gridded(
  data.in = data.mask,
  write.out = FALSE,
  shp.file = epu.shp,
  var.name = var.name,
  area.names = c("MAB", "GB"),
  start.time = 1,
  stop.time = 12,
  agg.time = "months",
  statistic = "mean"
)

# Plot the summer baseline climatology
terra::plot(data.clim[[1]], main = month.name[6:8])

```
## Step 5: Calculate Cell-by-Cell Spatial Anomalies

Finally, we calculate the spatial anomaly map. This is completed by subtracting the climatological `SpatRaster` from our summarized `SpatRaster` using `make_2d_anomaly_gridded`. Each pixel value in the resulting raster represents the departure from the summer climatology.


```{R}
data.anom <- make_2d_anomaly_gridded(
  data.in = data.summary,
  climatology = data.clim,
  shp.file = epu.shp,
  var.name = var.name,
  area.names = c("MAB", "GB"),
  write.out = FALSE
)

# Plot the anomaly map for the first processing period
terra::plot(data.anom[[1]])

```

## Step 6: Gridded Degree Days

For studies evaluating threshold-based biological stressors (such as spatial exposure to extreme bottom conditions), you can use `make_2d_deg_day_gridded_nc`. This function computes degree days, threshold counts, or consecutive threshold exceedances on a cell-by-cell basis.

Here, we calculate the gridded total number of days where Bottom Salinity was above `33` within the Georges Bank and Mid-Atlantic Bight regions.

```{R}
data.nd <- make_2d_deg_day_gridded(
  data.in = data.mask,
  write.out = FALSE,
  shp.file = epu.shp,
  var.name = "BottomS",
  statistic = "nd",
  ref.value = 33,
  type = "above",
  area.names = c("MAB", "GB")
)

# Plot the spatial threshold count grid
terra::plot(data.nd[[1]], main = "Gridded Days with Salinity > 33")

```