---
title: "2D Timeseries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2D Timeseries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Define parameters for cropping and subsetting
```{r setup}
library(here)
library(EDABUtilities)

epu.shp =system.file('data','EPU_NOESTUARIES.shp',package = 'EDABUtilities')
var.name = 'BottomS'
min.S = 30
max.S = 34

```

Crop the data based on the lat/lon bounds of epu.shp
```{r}

data.crop = crop_nc_2d(
  data.in = c(system.file('data','GLORYS_daily_BottomSalinity_2019.nc',package = 'EDABUtilities'),
                  system.file('data','GLORYS_daily_BottomSalinity_2020.nc',package = 'EDABUtilities')),
  shp.file = epu.shp,
  var.name = var.name,
  write.out = F,
  area.names =NA
)

terra::plot(data.crop[[2]])
```
Filter values within min.S to max.S range and remove values outside of the epu.shp regions
```{R}

data.mask = mask_nc_2d(
  data.in = data.crop,
  shp.file = epu.shp,
  min.value = min.S,
  max.value = max.S,
  binary = F,
  area.names = NA
)

terra::plot(data.mask[[1]])
```
Convert data.mask to data.frame aggregated by month and grouped by the regions MAB & GB

```{R}
data.ts = make_2d_summary_ts(data.in = data.mask,
                           shp.file = epu.shp,
                           var.name = var.name,
                           agg.time ='months',
                           statistic = 'mean',
                           area.names = c('MAB','GB'),
                           write.out = F)

ggplot2::ggplot(data.ts[[1]],ggplot2::aes(x = time, y= value))+
  ggplot2::geom_line()+
  ggplot2::facet_wrap(~area)

```
Create a monthly climatology from the raw ts data.frame. Filter data by summer months (6 through 8)
``` {R}
clim.ts =make_2d_climatology_ts(data.in =data.ts,
                            write.out = F,
                            start.time = 6,
                            stop.time = 8,
                            statistic = 'mean')

ggplot2::ggplot(clim.ts,ggplot2::aes(x=time,y = value, color = area))+
  ggplot2::geom_line()
  
```

Calculate the anomaly from original ts data and climatology
```{R}
anom.ts =make_2d_anomaly_ts(data.in = data.ts,
                   climatology = clim.ts,
                   write.out = F)

ggplot2::ggplot(dplyr::bind_rows(anom.ts),ggplot2::aes(x = time, y = anom.value, color = area))+
  ggplot2::geom_line()+
  ggplot2::facet_wrap(~ls.id, labeller = ggplot2::label_both)
```
Calculate the number of days above a reference salinity
```{r}
data.nday =make_2d_deg_day_ts(
  data.in = data.crop,
  shp.file = epu.shp,
  area.names =  c('MAB','GB'),
  var.name = 'BottomS',
  statistic = 'nd',
  ref.value = 33,
  type = 'above',
  write.out = F)

data.nday.all = dplyr::bind_rows(data.nday)
ggplot2::ggplot(data.nday.all,ggplot2::aes(x = ls.id, y = value, fill = area))+
  ggplot2::geom_bar(stat = 'identity', position = 'dodge')
  
```